# Transforming Data

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
```

## `{dplyr}` basics {-}

The book [R for Data Science](https://r4ds.had.co.nz/) is a *very* helpful reference guide. [Chapter 5](https://r4ds.had.co.nz/transform.html) covers many of the topics covered in this section, and may be useful as a resource later or to dive deeper into a topic. 

Topics:

- Filter rows
  - Comparisons
  - Logical operators
  - Missing values
- Arrange rows
- Select columns
- Add new variables 
- Grouped summaries
  - Combining multiple operations with the pipe
  - Missing values
  - Counts
  - Useful summary functions
  - Grouping by multiple variables
  - Ungrouping
- Grouped mutates (and filters)
- Extra: Working with factor variables


<!--
## Filter rows

- Comparisons
- Logical operators
- Missing values


## Arrange rows

## Select columns

## Add new variables 


##  Grouped summaries

- Combining multiple operations with the pipe
- Missing values
- Counts
- Useful summary functions
- Grouping by multiple variables
- Ungrouping


## Grouped mutates (and filters)



-->









## An example

The following example ties several of the above ideas together. Imagine that we have a set of grades from a course that we would like to convert to letter grades using a particular weighting and letter-grade cut-offs.

```{r}
url <- "https://raw.githubusercontent.com/haleyjeppson/R4QM/main/data/CourseData.txt"
students <- readr::read_table(url)[, -1]
students
head(students)
tail(students)
```

In this particular case we want the weighting to be:

- 20% to Hmwk (col 3)
- 25% to E1 (col 4)
- 25% to E2 (col 5)
- 30% to E3 (col 6)

But those weights could change later. A function can be written to take the data, and the weights and calculate the weighted average. We can then add that score to the students data frame, and add a final column with the actual letter grades on a 90-80-70-60 scale.

```{r}
students <- students %>% 
  mutate(Final = .2*Hmwk + .25*E1 + .25*E2 + .3*E3,
         Grade = case_when(
           Final < 60 ~ "F",
           Final < 69 ~ "D",
           Final < 79 ~ "C",
           Final < 89 ~ "B",
           Final >= 90 ~ "A"))

head(students)
```


<!-- ## Extra: Working with factor variables -->





# Reshaping Data

## `{tidyr}` basics {-}

The book [R for Data Science](https://r4ds.had.co.nz/) is a *very* helpful reference guide. [Chapter 12](https://r4ds.had.co.nz/tidy-data.html) covers many of the topics covered in this section, and may be useful as a resource later or to dive deeper into a topic. 

Topics:

- Tidy data
- Pivoting
  - Longer
  - Wider
- Separate & unite



## An example

For this example we begin with a subset of the 2009 [Program for International Student Assessment (PISA)](https://nces.ed.gov/surveys/pisa/) data. 

```{r}
## load data
url <- "https://raw.githubusercontent.com/haleyjeppson/R4QM/main/data/pisas09_subset.csv"
pisas <- readr::read_csv(url)
pisas
```

#### Pivot to long form {-}

Say we want to compare math and reading scores across a subset of countries. To do so, we will need to reshape our data into long form:

```{r}
## in long form
pisas_long <- pisas %>% 
  select(-sex) %>% 
  pivot_longer(cols = math:read, names_to = "item", values_to = "score")
pisas_long

pisas_long %>% 
  filter(country %in% c("Finland", "United States")) %>%
  ggplot(aes(x = score, weight = weight, fill = country)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~item) +
  theme_bw() +
  scale_fill_manual(values = c("#69cadf", "#f4602a", "#fbb03b", "#3273c2")[c(3,1)]) +
  labs(title = "Distribution of PISA 2009 Scores", x = NULL, y = NULL)
```

#### Pivot to wide form {-}

Say we want to plot male math scores against female math scores. To do so, we will need to summarize our data and then reshape our data into a wider format:

```{r}
## in wide form
pisas_wide <- pisas %>% 
  select(-read) %>% 
  group_by(country, sex) %>% 
  summarise(math = weighted.mean(math, weight, na.rm = TRUE)) %>% 
  pivot_wider(names_from = sex, values_from = math) 
  
pisas_wide

ggplot(pisas_wide, aes(x = Female, y = Male)) +
  geom_abline(alpha = 0.2) +
  geom_point() +
  labs(title = "PISA 2009 in Math - Gender Gap",
       x = "PISA 2009 Math score for girls",
       y = "PISA 2009 Math score for boys") +
  theme_light()
```

