# Manipulating Data

## dplyr basics {-}

## Filter rows

- Comparisons
- Logical operators
- Missing values


## Arrange rows

## Select columns

## Add new variables 


##  Grouped summaries

- Combining multiple operations with the pipe
- Missing values
- Counts
- Useful summary functions
- Grouping by multiple variables
- Ungrouping


## Grouped mutates (and filters)













## An example

The following example ties several of the above ideas together. Imagine that we have a set of grades from a course that we would like to convert to letter grades using a particular weighting and letter-grade cut-offs.

```{r}
students <- readr::read_table("data/CourseData.txt")[, -1]
head(students)
tail(students)
```

In this particular case we want the weighting to be:

- 20% to Hmwk (col 3)
- 25% to E1 (col 4)
- 25% to E2 (col 5)
- 30% to E3 (col 6)

But those weights could change later. A function can be written to take the data, and the weights and calculate the weighted average. We can then add that score to the students data frame, and add a final column with the actual letter grades on a 90-80-70-60 scale.

```{r}
students <- students %>% 
  mutate(Final = .2*Hmwk + .25*E1 + .25*E2 + .3*E3,
         Grade = case_when(
           Final < 60 ~ "F",
           Final < 69 ~ "D",
           Final < 79 ~ "C",
           Final < 89 ~ "B",
           Final >= 90 ~ "A"))

head(students)
```


## Extra: Working with factor variables




